Endogenous Politics and the Design of Trade Institutions\thanks{I thank Kyle Bagwell, Mostafa Beshkar, Rick Bond, Chad Bown, Nuno Limao, Giovanni Maggi, Anson Soderbery, Bob Staiger, Maurizio Zanardi, Ben Zissimos and participants in seminars at SITE for very helpful comments and discussions.}

### 2018-03-03 draft.tex
rewrite intro for submission to Midwest trade

\begin{document}
\title{Endogenous Politics and the Design of Trade Institutions\thanks{I thank Kyle Bagwell, Mostafa Beshkar, Rick Bond, Renee Bowen, Chad Bown, Paola Conconi, David DeRemer, James Lake, Nuno Lim\~ao, Giovanni Maggi, Devashish Mitra, Peter Rosendorff, Bob Staiger, Maurizio Zanardi, Ben Zissimos and participants in seminars at McGill, Rochester, Ryerson, Wesleyan, ETSG, InsTED, the Midwest International Trade Meetings, PEIO and SITE for very helpful comments and discussions.}}
\author{Kristy Buzard\thanks{Syracuse University, Economics Department. Email: kbuzard@syr.edu. http://faculty.maxwell.syr.edu/kbuzard.}}
\date{\vskip-.1in \today}
\maketitle

\vskip.3in
\begin{center} {\bf Abstract} \end{center}

\begin{quote}
{\small Much of the literature on the design of trade policy institutions takes the political pressure that governments face as exogenous. This paper integrates exogenous and endogenous political pressure into a single model, showing that the presence of endogenous political pressure overturns important results and provides insight into observed features of trade agreements as well as rules of organizations like the WTO. Endogenous politics destroys a traditionally-defined escape clause's ability to provide flexibility in the face of political shocks when lobbies use the flexibility to seek rents. This can explain why WTO Safeguard use is conditioned on measurable economic indicators.}
\end{quote}

@@

\bigskip
\section{Introduction}
\input{intro}

>>>>

\begin{document}
\title{Endogenous Politics and the Design of Trade Institutions\thanks{I thank Kyle Bagwell, Mostafa Beshkar, Rick Bond, Renee Bowen, Chad Bown, Paola Conconi, David DeRemer, James Lake, Nuno Lim\~ao, Giovanni Maggi, Devashish Mitra, Peter Rosendorff, Bob Staiger, Maurizio Zanardi, Ben Zissimos and participants in seminars at McGill, Rochester, Ryerson, Wesleyan, ETSG, InsTED, the Midwest International Trade Meetings, PEIO, SITE, and the University of Nebraska for very helpful comments and discussions.}}
\author{Kristy Buzard\thanks{Syracuse University, Economics Department. Email: kbuzard@syr.edu. http://faculty.maxwell.syr.edu/kbuzard.}}
\date{\vskip-.1in \today}

\maketitle

\vskip.3in
\begin{center} {\bf Abstract} \end{center}
\begin{quote}
{\small While most of the literature on the design of trade agreements and trading institutions takes the political pressure governments face to be exogenous, this paper endogenizes politics in a standard model for studying trade policy design questions. One can use this simple modeling framework to distinguish between the dynamics induced by exogenous political shocks and endogenous incentives of political actors, unifying these two strands of literature. The modeling framework can also provide fuller answers to trade policy design questions by elucidating the interactions between exogenous and endogenous political forces.  Applications to tariff caps and the escape clause show that important insights are missed if attention is restricted to exogenous political shocks. Most notably, endogenous politics destroys a traditionally-defined escape clause's ability to provide flexibility in the face of political shocks when lobbies use the flexibility to seek rents. This can explain why WTO Safeguard use is conditioned on measurable economic indicators.}
\end{quote}

@@

\bigskip
\section{Introduction}
The intensity and ubiquity of lobbying is a salient feature of trade policy making. The United States--Korea Free Trade Agreement was famously delayed by over four years, mainly by concerns from the U.S. auto industry. Over the last two decades in the United States alone, anywhere between 700 and 1200 entities reported lobbying on trade issues each year. \Textcite{bt} find that total annual lobbying expenditure on trade issues in 1999-2001 when Lobbying Disclosure Act data first became available was about $\$$200 million. Campaign contributions are harder to attribute cleanly to a specific issue but are likely an order of magnitude larger.

Despite the obvious importance of interest groups in the trade policy-making process, relatively few papers have examined lobbying incentives in trade agreement and trade institution design. This is understandable, as models of endogenous politics following \Textcite{gh94} are complex. It can be quite challenging to endogenize politics along these lines while also taking account of complex institutional features. Thus we have very important (e.g. \Textcite{mitra}, \Textcite{mrc2007}) but also very few models in this literature in which political pressure is modeled as endogenous.

However, it seems likely that many questions of trade agreement and trade institution design are impacted in important ways by decisions taken by political pressure groups, who themselves respond to features of the designed environment. In order to take account of endogenous politics when asking questions of trade policy design, I use a simple model that is standard in the literature (e.g. \Textcite{bs2001}, \Textcite{bown2002}, \Textcite{bown2004}, \Textcite{bs2005}, \Textcite{martinvergote}, \Textcite{bagwell2009}, \Textcite{beshkar2010a}, \Textcite{ms2012a}, and \Textcite{ms2013}) and add an endogenously-determined element to the usual exogenously-determined political economy weights. This framework is more tractable than those in the \Textcite{gh94} tradition and allows the incorporation of endogenous politics when asking a rich set of institutional design questions, including in repeated-game settings.\footnote{Repeated non-cooperative-game models of trade agreements have been considered by \Textcite{mcm86,mcm89}, \Textcite{dixit1987}, \Textcite{bs1990, bs1997a, bs1997b, bs2002}, \Textcite{kovthurs}, \Textcite{maggi99}, \Textcite{coatesludema}, \Textcite{ederington}, \Textcite{rosendorff}, \Textcite{bagwell2009}, and \Textcite{park}. \Textcite{buzard2013a} features a repeated-game model similar in spirit to the model under consideration here but focusing on questions of optimal punishments when the government is non-unitary.}

After introducing the modeling framework, I use it to carefully examine the assumptions implicit in the government objective functions that are most commonly used in the literature. The \Textcite{baldwin}-style  government objective function is everywhere increasing in political pressure. \Textcite{gh94} and its extension to trade agreements in \Textcite{gh95} provide microfoundations for this \Textcite{baldwin}-style government objective function, where a fixed weight of $1+a$ is attached to the surplus of groups who lobby, and $a$ to the surplus of those who don't lobby.\footnote{Note that $a$ is the weight the government places on social welfare.} This cannot microfound a flexible model as in \Textcite{longvousden} or in which shocks to the political-economy weights come from shocks to any parameter other than the weight the government places on social welfare.

More importantly, this standard functional form leads inexorably to the conclusion that governments do not want to discourage political pressure except in the presence of some kind of dynamic inefficiency such as in \Textcite{mitra} or \Textcite{mrc2007}. The modification to the government objective function I propose leads to the result that governments may indeed want to use trade agreements to reduce lobbying in a much wider set of circumstances than those identified in the previous literature.\footnote{With only one lobby in each country, the governments will be unable to encourage lobbying in excess of that which is optimal from the lobby's point of view.} Examining this alternative welfare function in combination with endogenous lobbying can provide a bridge between the theoretical literature and the claims of trade policy practitioners that an important role of trade agreements is to rein in protectionist pressure.

Having looked carefully at the government's objective function, I then use the model to carefully distinguish between the dynamics induced by exogenous and endogenous politics for two example policies: tariff caps and escape clauses. I show that the model captures results from both the exogenous politics (i.e. \Textcite{bs2005} and the endogenous politics (i.e. \Textcite{mrc2007}) literatures on tariff caps. The presence of endogenous political pressure changes the optimal choice of tariff bindings within a trade agreement in quite significant ways compared with the case where political pressure is assumed to be exogenous. Additionally, endogenous political pressure makes it harder to incentivize cooperation through repeated game incentives. As in \Textcite{mrc2007}, I show in this simpler framework that allows for easy comparison to the exogenous shocks case that one of the uses of tariff caps is to incentivize the lobby to engage in the political process after the trade agreement is in place. Combined with the idea that governments can use tariff caps to restrain endogenous political pressure, a story emerges in which governments employ trade agreements to carefully manipulate lobbying incentives in order to maximize their political objectives.

The changes induced by endogenous politics are even more stark for the escape clause. Here one finds that the standard, theoretical escape clause can't work in the presence of endogenous political pressure. The idea of the escape clause is that it is often advantageous to grant governments a period of relief from trade commitments when there is asymmetric information about the strength of ex-post political economy shocks. That is, one would rather allow a short period of ``escape'' from the agreement rather than have the agreement violated because domestic political opposition is temporarily too strong to be resisted. Unfortunately, the logic breaks down in the presence of endogenous political pressure. An escape clause allows a government to apply a higher tariff barrier when it experiences intense political pressure. When political pressure is chosen endogenously by special interest groups \textit{and} the government gets a free pass at the WTO whenever it feels sufficient political pressure from domestic interest groups, those interest groups have a strong incentive to exert the required level of pressure regardless of the underlying state of the world, eviscerating the escape clause. I argue on this basis that if design choices might affect lobbying incentives, one should carefully consider when it is appropriate to assume that political pressure is given exogenously.

When we turn to another attractive feature of the model---that one can study the simultaneous effects and interactions of exogenous and endogenous political forces---we derive a clear explanation for why the conditions for invoking the WTO Safeguards measure rely on purely observable economic variables, and why the level of protection governments can choose when invoking a Safeguard is related to the injury imposed by the shock and not its political impact. The model formalizes the intuition that if the WTO's Dispute Settlement Body actually followed the procedures suggested by the literature---that is, it enforced commitments based on signals of the political pressure experienced by governments---pressure groups would simply exert more pressure and the Safeguard clause could not play the escape-providing role for which it was intended. On the other hand, if the purpose of contingent protection measures is actually to provide a political safety valve, this analysis could help explain the infrequency with which safeguards are invoked under the WTO, as I argue below that the WTO's escape clause appears to have been designed to provide relief only from economic shocks.

We miss important dynamics like this one when we assume political pressure is exogenous as is done for tractability in almost all the literature that studies trading rules and institutions (see \Textcite{mrc2007}, \Textcite{lt} and \Textcite{buzard2013a} for exceptions).\footnote{\Textcite{clz} also endogenize lobbying using a new approach, that of a contest model.} The main contribution of this paper is to provide a tractable framework for integrating lobbying incentives into these analyses. While others (cfr. \Textcite{mrc2007}, \Textcite{mitra}) have formulated models in which some dynamic inefficiency induces a government objective that is non-monotonic in lobbying effort, I can achieve results with the same flavor in a more tractable framework. The framework can therefore be used to study a wider range of questions, including dispute settlement design, optimal retaliation schemes and the depth of integration among others. It can also examine exogenous shocks to political weights and lobbying incentives simultaneously, unifying the two distinct literatures.

Through the lens of the examples considered in this paper, one can see that modeling choices concerning both the source of political pressure and the form of governments' political objective functions have important impacts on questions of the design of and motives behind trade agreements as well as the institutions that help to enforce them. It is important to consider how these features interact with the question being asked. For some inquiries, such as those that are the focus of \Textcite{ms2011, ms2012a} in which the impact of the political economy parameter on the government's objective function is not crucial, there may be no loss in employing a simple form with exogenous political pressure. However, for other questions in which lobbying incentives might be impacted or the government's desire for a commitment mechanism is central, a more general treatment seems prudent.

\textbf{A few additional comments on the literature}. Tariff caps have received particular attention in the literature that takes political pressure to be exogenous. The basic result on the political efficiency of tariff caps established in \Textcite{bs2005} has been extended in numerous directions. \Textcite{hms} show that the basic logic concerning tariff caps is unaltered in a model with contracting costs and multiple policy instruments, while \Textcite{ls} show that the result holds when fines are allowed as punishments, although the politically efficient tariffs may not be achievable when enforcement of the punishment is taken into consideration.

\Textcite{ab2012} consider that the weight that tariff revenue receives in the government welfare function may be private information and show that tariff caps remain optimal under certain conditions, while \Textcite{ab2013} show that the result goes through when the \Textcite{bs2005} model is generalized to include monopolistic competition as well as more general payoff and distribution function. \Textcite{bb} extend the theory to an environment with asymmetric country sizes with costly state verification and show that tariff caps and escape clauses can both emerge endogenously in optimal trade agreements. \Textcite{bbr} and \Textcite{nos} provide evidence on the relationship between tariff caps, binding overhang and market power that confirms the predictions of the terms of trade theory.

On escape clauses, \Textcite{bs1990} is similar in spirit to \Textcite{bs2005}. Here, there is no private information and trade volume shocks in the presence of self-enforcement constraints are behind the need for an escape clause. \Textcite{bc} find empirical support for this theory. \Textcite{beshkar2010b} compares the GATT escape clause to the WTO Safeguards agreement and shows that the Dispute Settlement Body as a non-binding arbitrator can assist governments in self-enforcing their trade agreements. \Textcite{ms2012a} study the optimal form of the trade agreement in the presence of renegotiation where compensation between governments is inefficient.

In the next section, I present the model and an in-depth discussion of the political welfare function of the government. Section~\ref{sec:rigid} contains an analysis of tariff caps without the flexibility of escape and introduces the repeated-game environment. Section~\ref{sec:escape} explores questions regarding the design of the escape clause. Section~\ref{sec:concl} concludes.

%\input{intro}
